---
title: "R Workshop for Applied Research Projects"
subtitle: "A practical bridge for Stata users"
author: "J. Montornes"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)
library(dplyr)
library(ggplot2)
library(fixest)
library(lubridate)
library(stringr)
theme_set(theme_minimal(base_size = 14))
sample_data <- read.csv("data/sample_economy.csv")
```

## Workshop goals

- Understand the key differences between R and Stata.
- Write reproducible scripts instead of isolated console commands.
- Manipulate data with `dplyr`.
- Produce charts, regressions and simple output tables.
- Know where to find help and how to structure a small project.

## Agenda

1. Interface, help and project structure.
2. Packages, objects and syntax.
3. Data import and data manipulation.
4. Charts and tables.
5. Econometric estimation.
6. Dates, strings and wrap-up exercise.

## RStudio interface

- **Script**: saved, versionable and rerunnable source code.
- **Console**: interactive execution and error messages.
- **Environment**: objects currently stored in memory.
- **Files / Plots / Packages / Help**: outputs, files and documentation.

Useful habit: anything you want to keep should be written in a script, not only typed in the console.

## Help and documentation

```{r eval=FALSE}
?lm
help("merge")
example(mean)
```

Useful resources:

- [R for Data Science](https://r4ds.hadley.nz/)
- [Stata2R](https://stata2r.github.io/)
- [UtilitR](https://www.utilitr.org/)

## Packages

A package adds functions, datasets or methods.

```{r eval=FALSE}
install.packages("dplyr")  # once per machine
library(dplyr)             # in each new session
```

For training, listing required packages in the README or an install script prevents many last-minute issues.

## Objects and assignment

```{r}
baguette_price <- 0.95
quantities <- c(1, 2, 3)
spending <- baguette_price * quantities
spending
```

In R, results are usually assigned to objects with `<-`. The `=` sign often works, but `<-` remains the clearest convention.

## Useful differences with Stata

| Stata | R |
|---|---|
| One active dataset in memory | Several objects at the same time |
| `use`, `save` | `read.csv()`, `write.csv()`, `readxl::read_xlsx()` |
| `gen`, `replace` | `mutate()` |
| `keep if` | `filter()` |
| `collapse` | `group_by()` then `summarise()` |
| `merge` | `left_join()`, `inner_join()`, `full_join()` |

## Importing a file

```{r}
economy <- read.csv("data/sample_economy.csv")
head(economy)
```

For Excel files:

```{r eval=FALSE}
readxl::read_xlsx("data/my_file.xlsx", sheet = "Sheet1")
```

## Selecting and filtering

### Stata

```stata
keep if country == "France" & year >= 2021
keep country year sector wage
```

### R

```{r}
economy |>
  filter(country == "France", year >= 2021) |>
  select(country, year, sector, wage)
```

## Creating a variable

### Stata

```stata
gen real_wage = wage / price_index * 100
```

### R

```{r}
economy <- economy |>
  mutate(real_wage = wage / price_index * 100)

economy |>
  select(country, year, sector, real_wage) |>
  head()
```

## Aggregating

### Stata

```stata
collapse (mean) mean_wage = wage, by(country year)
```

### R

```{r}
country_year <- economy |>
  group_by(country, year) |>
  summarise(mean_wage = mean(wage), .groups = "drop")

country_year
```

## Joining two tables

```{r}
population <- data.frame(
  country = c("France", "Germany", "Italy"),
  population_m = c(67.8, 83.2, 59.0)
)

country_year |>
  left_join(population, by = "country") |>
  head()
```

Remember the direction of the join: `left_join(x, y)` keeps all rows from `x`.

## data.table in one sentence

`data.table` is very efficient on large datasets. Its syntax is compact:

```{r}
library(data.table)
dt <- as.data.table(economy)
dt[country == "France", .(mean_wage = mean(wage)), by = year]
```

For beginners, `dplyr` is often easier to read; for massive datasets, `data.table` may become preferable.

## Simple chart

```{r}
ggplot(country_year, aes(x = year, y = mean_wage, color = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(x = NULL, y = "Average wage", color = NULL)
```

## ggplot2 principles

- `ggplot(data, aes(...))` declares the data and mapped variables.
- `geom_*()` chooses the visual representation: points, lines, bars.
- `labs()`, `scale_*()` and `theme_*()` adjust presentation.
- Each layer is added with `+`.

```{r eval=FALSE}
ggplot(data, aes(x = x, y = y)) +
  geom_point() +
  labs(x = "X", y = "Y")
```

## Linear regression

### Stata

```stata
reg real_wage employment i.sector
```

### Base R

```{r}
model_lm <- lm(real_wage ~ employment + factor(sector), data = economy)
summary(model_lm)
```

## Fixed effects with fixest

`fixest` is close in spirit to `reghdfe`: fast, concise and convenient for fixed effects and clustered standard errors.

```{r}
model_fe <- feols(
  real_wage ~ employment | country + year,
  data = economy,
  vcov = ~ country
)
etable(model_fe)
```

## Clean output

```{r}
broom::tidy(model_lm)
```

For a report:

- keep model objects (`model_lm`, `model_fe`);
- generate tables in the script;
- avoid manual copy-paste from the console.

## Dates with lubridate

```{r}
day <- ymd("2023-02-11")
day
year(day)
month(day)
```

`ymd()`, `dmy()` and `mdy()` make the day/month/year order explicit.

## Strings with stringr

```{r}
text <- "Hello world"
str_replace(text, "world", "universe")
str_sub(text, 1, 5)
str_detect(text, "Hello")
```

Mental translation:

- Stata `subinstr()` -> R `str_replace()` or `str_replace_all()`.
- Stata `substr()` -> R `str_sub()`.
- Stata `regexm()` -> R `str_detect()`.

## Web calls and external databases

For DB.NOMICS or other APIs, showing non-executed code keeps slides robust.

```{r eval=FALSE}
library(rdbnomics)
df <- rdbnomics::rdb(ids = "INSEE/CNT-2014-CSI/...")
```

In training, always provide a local fallback file in case the network, proxy or API is unavailable.

## Recommended mini-project

1. Import `data/sample_economy.csv`.
2. Create a real wage variable.
3. Compute average wages by country and year.
4. Plot the time series.
5. Estimate a simple model with fixed effects.
6. Save the chart and the result table.

## Project good practices

```text
AtelierR/
  data/
  exercises/
  solutions/
  atelierR.Rmd
  atelierR_en.Rmd
  README.md
```

- Use relative paths from the project root.
- Store lightweight example data in `data/`.
- Keep executable solutions in `solutions/`.
- Document goals, prerequisites and rendering commands in the README.

## Conclusion

R is not only a statistical calculator: it is a project environment.

The main gain over purely interactive work is reproducibility: tomorrow, someone should be able to rerun the script and obtain the same outputs.
